---
title: "WAFishBiology_vignette"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{WAFishBiology_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r library_setup}
library(WAFishBiology)
Fig_Path = "C:/~/WAFishBiology/vignettes/"
```


# Introduction 
#### Last updated May 2023 (Alex Hesp - DPIRD)

This package is intended as a toolbox of methods for describing various biological traits of fish and invertebrate species, including growth, size/age at maturity, weight-length relationships etc., which represent key information for many model-based fisheries assessments. The various R functions output a range of statistics and diagnostic plots. The various biological relationships and associated parameters are required inputs a range of assessment models, e.g. per recruit analysis and statistical catch at age (and length) models.

Please note that this is a 'trial version' with testing of the various routines continuing. Please report any identified issues to Alex Hesp asap $(\mathrm{Alex.Hesp@dpird.wa.gov.au})$.

## Growth curves 
Several functions are available that relate to fitting growth curves to length-at-age data or data derived from length mixture distribution analysis. The alternative growth models include the von Bertalanffy growth model (von Bertalanffy, 1938), the generalised growth model of Schnute (1981) and the seasonal growth curve of Somers (1988). The routines implemented allow for either a single curve to be fitted to the data, or two curves (i.e. females and males, or area 1 and area 2) to be fitted simultaneously. In the latter case, common variance is assumed for the two curves.

### von Bertalanffy growth model
Fitting a single von Bertalanffy growth curve to data, $L_t$, the estimated mean length of fish at age $t$, is

$$
L_t = L_\infty(1-\exp(-k(t-t_0)))
$$
where $L_\infty$ is the asymptotic length, $k$ is the growth coefficient and $t_0$ is the hypothetical age at zero length.

### Schnute growth model

Schnute's growth model includes four parameters, including $a$ (1/time), $b$ (dimensionless) and two sizes $L_1$ and $L_2$, corresponding to references ages $\tau_1$ and $\tau_2$, respectively. Depending on whether or not $a$ and/or $b$ are equal to zero, the Schnute growth functions takes one of the following four forms: 

$a$ <> 0, $b$ <> 0
$$
L_t =\biggl[{L_1}^b+({L_2}^b-{L_1}^b)\frac{(1-\exp(-a(t-\tau_1)))}{(1-\exp(-a(\tau_2-\tau_1))}\biggr]^\frac{1}{b}
$$
$a$ <> 0, $b$ = 0 
$$
L_t ={L_1}\biggl[\ln(L_2/L_1)\frac{(1-\exp(-a(t-\tau_1)))}{(1-\exp(-a(\tau_2-\tau_1))}\biggr]
$$
$a$ = 0, $b$ <> 0 
$$
L_t =\biggl[{L_1}^b+({L_2}^b-{L_1}^b)\frac{(t-\tau_1)}{(\tau_2-\tau_1)}\biggr]^\frac{1}{b}
$$
$a$ = 0, $b$ = 0
$$
L_t ={L_1}\biggl[\ln(L_2/L_1)\frac{(t-\tau_1)}{(\tau_2-\tau_1)}\biggr]^\frac{1}{b}
$$

### Somers (1988) seasonal growth curve

Somers (1988) seasonal growth curve is
$$
L_t = L_\infty(1-\exp(-k(t-t_0)))-S(t)+S(t_0)
$$
where
$$
S(t)=\frac{CK}{2\pi} \sin{2\pi}(t-t_s)
$$
$$
S(t_0)=\frac{CK}{2\pi} \sin{2\pi}(t_0-t_s)
$$
In the above equations, $C$ modifies the amplitude of the growth oscillations and $t_s$ represents the time between time 0 and the start of the convex portion of the first sinusoidal growth oscillation, i.e. the inflection point (see Garcia-Berthou et al., 2012). If C = 0, then there is no seasonal oscillation and the model reduces to the typical von Bertalanffy growth model. If C = 1, then growth completely ceases during one part of the year.

### Objective function for fitting growth curves to length-at-age data

The objective function minimised by nlminb is the negative log-likelihood, calculated as
$$
\lambda=\frac{n}{2}\bigg[\ln(2\pi)+2\ln(\hat{\sigma})+1\bigg]
$$
see (e.g. Haddon, 2011). Non-linear regression (employing the optimisation routine nlminb) is used to fit the growth curve, given a set of age and length data, and a set of starting values for the von Bertalanffy growth parameters (Linf, vbK, and tzero):  

### Objective function for fitting growth curves to data derived from mixture distribution analysis

The objective function minimised by nlminb is the negative log-likelihood, calculated as

$$
\lambda=0.5\log_e(\sigma_M^2+\sigma_P^2)+0.5\log_e(2\pi)+\frac{(X-\mu)^2}{2(\sigma_M^2+\sigma_P^2)}
$$
where $X$ is the “observed” mean size-at-age calculated from the mixture analysis, $\sigma_M^2$ is the variance associated with the “observed” mean size-at-age from the mixture analysis, $\sigma_P^2$ is an
additional process error variance estimated in the growth curve analysis, and $\mu$ is the estimated mean size-at-age, according to the growth curve.

## Mixture analyses (using fish length data)

The model allows normal or gamma mixture distributions to be fitted to length composition data with up to three modes (assumed to represent different fish cohorts). For a given length sample with a single mode, if it can be assumed that these data conform to a normal distribution, the expected relative proportion of fish in each size class is calculated as 

$$
\hat{P}_{S_i} = \int_{S_i^-}^{\mathrm{S_i^+}}\frac{1}{\sigma\sqrt{2{\pi}}}\exp\bigg[\frac{(l-\mu^2)}{2\sigma{^2}}\bigg]dl
$$
where $l$ is the length of the fish, $S_i^-$ and $S_i^+$ are the lower and upper bounds of size class $S_i$, respectively, and $\mu$ and $\sigma$ are the mean and standard deviation, respectively, of the normal distribution describing the size distribution (Marks et al., 2020). If the data are assumed to conform to a gamma distribution, $\hat{P}_{S_i}$ is calculated as

$$
\hat{P}_{S_i} = \int_{S_i^-}^{\mathrm{S_i^+}}\frac{1}{\beta^\alpha\Gamma(\alpha)}l^{\alpha-1}\exp(-l/\beta){dl}
$$
where where $\alpha$ and $\beta$ are the ‘scale’ and ‘shape’ parameters, respectively, of the gamma distribution describing the size data 

When two modes are present (assumed to represent 2 age cohorts), the size frequency data is assumed to represent a random sample from a multinomial distribution, with the overall size distribution being comprised of two normal distributions. Using $0^+$ and $1^+$ to represent the values of the smaller and larger cohort, respectively, $\hat{f}_{S_i}$, the overall expected frequency of fish in size class $S_i$ in the sample is calculated as

$$
\hat{f}_{S_i} = n\bigg[\hat{P}_{S_i,0+}\phi_{0+}+\hat{P}_{S_i,1+}(1-\phi_{0+})\bigg]
$$
where $n$ is the sample size for the month and $\phi_{0+}$ is the proportion of fish belonging to the 0+ cohort (estimated as a model parameter). Where three cohorts are present, $\hat{f}_{S_i}$ is calculated as 

$$
\hat{f}_{S_i} = n\bigg[\hat{P}_{S_i,0+}\phi_{0+}+\hat{P}_{S_i,1+}\phi_{1+}+\hat{P}_{S_i,2+}(1-\phi_{0+}-\phi_{1_+})\bigg]
$$
The expected relative proportion of fish in size class $S_i$ is be calculated as

$$
\hat{P_{S_i}}=\hat{f_{S_i}}/\sum_{S_i}\hat{f_{S_i}}
$$
The mixture distribution model is fitted to the sample data by minimizing the negative log-likelihood, calculated as

$$
\lambda=\sum{S_i}{f_{S_i}}\log_e(\hat{P_{S_i}})
$$
where $f_{S_i}$ is the observed frequency of fish in size class $S_i$. Estimates of asymptotic standard errors (used to calculate approximate 95% confidence limits) for the estimated model parameters are derived from a variance-covariance matrix, calculated from the Hessian matrix outputted by nlminb in R. The estimated model parameter include mean and standard deviation for each cohort, and also proportion of the catch in the 0+ cohort (when two cohorts are present) or proportions in the 1+ and 2+ cohorts (when three cohorts are present).

## Available functions related to fish growth analysis using length or length-at-age data

Available functions for visualising length-at-age data, and visualising growth curves fitted to these data are:  
SimulateLengthAtAgeData_vonBert()  
SimulateLengthAtAgeData_Schnute()  
SimulateLengthAtAgeData_SomersSeasonal()  
PlotLengthAtAgeData()  
PlotFittedGrowthCurve()   
CalcLengthAtAge_SomersSeasonalGrowthCurve()  
SchnuteGrowthfunction()  

The available functions for fitting growth curves to length-at-age data are:    
GetvonBertalanffyGrowthResults()  
GetSchnuteGrowthResults()  
GetSeasonalGrowthResults()  

The available functions related to mixture distribution analyses are:  
GetMixtureModelResults()  
PlotMixtureDistnResults()  
 

## Growth analysis from tagging data

Four models are available in this R package for describing growth from tagging data. As described in Strain et al. (2017), these include the von Bertalanffy growth model (assuming a linear pattern of change in growth rate with respect to initial length), the Gompertz growth model (allows for initially increasing and then decreasing growth rates as the initial lengths of animals increase), a Gaussian function (allowing for low growth rates over a range of both small and large initial lengths) and a double-logistic curve (which incorporates the “inverse logistic” curve described by Haddon et al. (2008) but also contains an ascending logistic curve for describing the growth of the very small individuals).

### von Bertalanffy growth model - tag increment data

Applying the von Bertalanffy growth model, $dL_t/dt$, the instantaneous rate of change in 
length of an individual, given its initial length at age $t$ years, $L_t$, is

$$
dL_t/dt=(L_{\infty}-L_t)(1-\exp(-k\Delta_t))
$$
where $L_\infty$ is the average maximum length of individuals in the population, $k$ is the growth 
coefficient, and $\Delta_t$ is a minute number between capture and recapture (e.g. Haddon, 2011). 

### Gompertz growth model - tag increment data

Fitting the Gompertz growth model, $dL_t/dt$ is estimated as

$$
dL_t/dt=L_{\infty}\bigg(\frac{L_t}{L_{\infty}}\bigg)\exp(-g\Delta_t)
$$
where $g$ is a growth constant (Heliodoniotis et al. 2011). 

### Gaussian function - tag increment data

Using the Gaussian function, $dL_t/dt$
is estimated as 

$$
dL_t/dt=A\exp\bigg(\frac{-(L_t-\mu)^2}{2\sigma^2_{GF}}\bigg)
$$
where $A$ is the maximum growth (mm $y^{-1}$), $\mu$ is the size at maximum growth (mm) and $\sigma_{GF}$ is the standard deviation of the distribution of maximum growth vs size (Rogers-Bennett et al. 2007). 

### Double logistic curve - tag increment data

Using the double-logistic curve, $dL_t/dt$ is

$$
dL_t/dt=a\Bigg(1/\bigg[1+\exp(-\log_e(19)(L_t-L_{{50}_1})(L_{{95}_1}-L_{{50}_1})\bigg]\Bigg)
\Bigg(1/\bigg[1+\exp(\log_e(19)(L_t-L_{{50}_2})(L_{{95}_2}-L_{{50}_2})\bigg]\Bigg)
$$
where $a$ is the estimated maximum length increment attained by individuals at any initial length, $L_{{50}_1}$ is the initial length midway between the smallest and maximum growth increment on the ascending limb of the curve, $L_{{95}_1}$ is the initial length at which 95% of the difference between the smallest growth increment and maximum growth increment on the ascending limb of the curve is reached, $L_{{50}_2}$ is the initial length midway between the largest and smallest growth increment on the descending limb of the curve, and $L_{{95}_2}$ is the initial length at which 95% of the difference between the largest and smallest growth increment on the descending limb of the curve is reached. 

### Inverse logistic curve - tag increment data

The “inverse logistic” curve (e.g. Haddon et al., 2008) is as described above for the double logistic model, but with only a descending limb, i.e.

$$
dL_t/dt=a\Bigg(1/\bigg[1+\exp(-\log_e(19)(L_t-L_{{50}_1})(L_{{95}_1}-L_{{50}_1})\bigg]\Bigg)
$$
    
### Exponential decay curve - tag increment data

$$
dL_t/dt=a\exp(-L_t/\tau)
$$
where $a$ is the maximum length increment attained by individuals at any initial length and $\tau$ is the rate of change in growth with length. This model is similar to that described by deLestang (2018), but with one less parameter, i.e. excluding a term describing the minimum length increment attained by individuals.

### Calculation of expected lengths for specified time intervals (with any of the above tagging models)

Applying the Gaussian function, for example, if the length of a fish at time $t=t_0$ is $L_{t=t_0}$, then its expected length $\hat{L}_{t=t_0+T}$ at time $t=t_0+T$ is

$$
\hat{L}_{t=t_0+T}=\hat{L}_{t=t_0}+\Delta{L_{t_0,t_{0+T}}}=L_{t=t_0}+\int_{t=t_0}^{t=t_0+T}A\exp\bigg(\frac{-(L_t-\mu)^2}{2\sigma^2_{GF}}\bigg)dt
$$
where $\hat{L}_{t=t_0+T}$ is the expected length increment between $t=t_0$ and $t=t_0+T$. For each 
tagged individual, this “initial value problem” is solved using a 4th order Runge-Kutta algorithm to calculate an estimate of its expected length at its time of re-measurement given the values of the growth parameters $A$, $\mu$ and $\sigma^2_{GF}$. The parameters of the growth model  (as described in this example by the above Gaussian function) are estimated by minimising the negative  log-likelihood where it is assumed that, for individual $j$, the observed length increment $X_j$ represents a random variable drawn from the normal distribution about the estimated increment $\hat{X}_j$, given its initial length and the time between its tagging and re-measurement. The overall log-likelihood for the growth curve is calculated as 
 
$$
\lambda=-\frac{n}{2}\bigg[\log_e(2\pi)+2\log_e(\hat{\sigma})+1\bigg]
$$

where $n$ is the number of animals for which measurements were taken of their initial and final length lengths and $\hat{\sigma}$ is the standard deviation associated with the deviations between the observed and predicted final lengths.

### Available functions related to fish growth analysis using length or length-at-age data

Available functions for visualising tagging data, and visualising growth curves fitted to these data are: 
GetTaggingGrowthModelResults()
PlotFittedTaggingGrowthModelResults(()
SimulateTagRecaptureData()

## Analysis of fish weight-length data

The relationship between the natural logarithms of fish weight $(W)$ and fish length $(L)$ may be described by a linear model, i.e.

$$
\log_eW=\log_e(a)+b\log_eL
$$

where $a$ and $b$ are the parameters of the model. In this R package, the model is fitted in R using the lm() function, and the predict() function is use to calculate 95% confidence limits (associated with the line) and 95% prediction intervals. For plotting, the data may be back log-transformed. The back-transformed values are weight are corrected for bias, associated with back-transformation, calculated as

$$
W=\exp(\log_eW)\exp(0.5\sigma^2)
$$
where $\sigma^2$ is the residual variance, calculated when fitting the linear regression model to the log-transformed length and weight data.

### Available functions related to analysis of fish weight and length data

Available functions for visualising fish weight and length data, and visualising curves fitted to these data are: 
GetWeightLengthRegressionResults()
PlotWeightLengthRel_LogSpace()
PlotWeightLengthRel_NormalSpace()
SimulateWeightLengthData()

## Reproductive analyses 

A logistic function may be used to describe the probability of a fish being mature at a given length (or age). The routines implemented allow for either a single curve to be fitted to the data, or two curves (i.e. females and males, or area 1 and area 2) to be fitted simultaneously. In the latter case, common variance is assumed for the two curves. Several alternative logistic functions are available, including models with fixed height (at 1), variable height, symmetric curves and asymmetric curves.  

### Logistic size and age at maturity (or sex change) curves
The probability, $P$, of a fish, of length $L$ being mature, may be described using the following symmetric logistic function
$$
\rho = {P_{\max}/(1+\exp(-\log_e(19)(L-L_{50})/(L_{95}-L_{50})))}
$$
where P_{max} is the maximum proportion of fish that are mature, $L_{50}$ and $L_{95}$ are the lengths at which 50 and 95% of fish are mature, respectively. The same form of equation may be used to describe the probability of maturity of fish at age, replacing $L$ with fish age, and $L_{50}$ and $L_{95}$ with $A_{50}$ and $A_{95}$, i.e. the ages at which 50 and 95% of fish are mature, respectively. In cases where $P_{\max}$ can be assumed to be equal to 1, only two parameters (either $L_{50}$ and $L_{95}$, or $A_{50}$ and $A_{95}$) are input into the model. Alternatively, the following, more flexible function, allowing for asymmetry, may be applied

$$
\rho = {P_{\max}/((1+Q\exp(-BL))^{1/V})}
$$
where $P_{max}$, $Q$, $B$ and $v$ are estimated parameters. Similar to above, where P_{max} can be assumed to equal one, only 3 parameters need to be estimated.


### Objective function for fitting size or age at maturity (or sex change) curves

The negative log-likelihood associated with the above logistic curve may be calculated as

$$
\lambda = \sum_{j}X_j[\ln\rho_j+(1-X_j)\ln(1-\rho_j)]
$$
where the $j^{th}$ fish was represented by $X=0$ if it was immature and $X=1$ if it was mature (e.g. French et al., 2014). The same form of equation may be used to describe the probability of a fish being mature given its age. Likewise, this relationship may also be used to describe the process of sex change for hermaphroditic species (by length or age), for cases where all fish undergo sex change if they survive long enough to reach the largest size class (or oldest age class).

### Available functions related to analysis of fish length at maturity and age at maturity (or sex change) data

Available functions for visualising fish length at maturity and age at maturity data, and visualising curves fitted to these data

The available functions for visualising size and age at maturity data, and fitting curves to these data are:  
CalcPropMatureAtLength()  
CalcPropMatureAtAge()  
GetLogisticMaturityCurveResults()  
PlotAgeAtMaturityData()  
PlotFittedAgeAtMaturityCurve()  
PlotLengthAtMaturityData()  
PlotFittedLengthAtMaturityCurve()  
SimulateLengthAtMaturityData()  
SimulateAgeAtMaturityData()  

### Available functions related to mean monthly gonadosomatic indices GSIs)

Routines are available for rapid calculation and plotting of mean monthly gonadosomatic indices (GSIs) and associated 95 percent confidence limits. As standard in fish biological studies, GSIs for individual fish are calculated as

$$
GSI = \bigg(\frac{GW}{BW}\bigg)100
$$
where $GW$ is the gonad weight and $BW$ is the total body weight of the fish. 95% confidence limits for mean monthly gonad weights are calculated as 1.96 multiplied by their standard errors.

Available functions include:  
CalcMeanMonthlyGSIs()  
PlotMeanMonthlyGSIs()  

### Available functions related to mean monthly standardised gonad weights

Trends in mean monthly GSIs can be impacted by variations in the sizes of fish collected in monthly samples. Thus, some researchers prefer to consider trends in mean monthly gonad weights (or GSIs) after they have been standardised for length effects, such as by using Analyis of Covariance (ANCOVA) linear model or generalised linear model (e.g. Cossington et al., 2011). WAFishBiology calculates standardised mean monthly gonad weights (and associated 95% confidence limits), using the linear model (lm) function within R. The model used to calculate mean gonad weights according to fish total length and month is

$$
\log_e{GW}_{i,k}=M_k+a{\log_e{L_i}}{\sim} N(0,\sigma^2)
$$
where $GW_{i,k}$ is the standardised gonad weight, $M_k$ is the month effect and $L_i$ is the length effect. 

Available functions include:  
CalcMeanMonthlyStGonadWts()  
PlotMeanMonthyStGonadWts()

### Available functions related to monthly prevalences of fish at different gonadal development stages

Some routines have been added for calculation and plotting of monthly prevalences of fish at different gonadal development stages. Available functions include:  

CalcMonthlyGonadStageProps()  
PlotMonthlyGonadStageProps()  

### Spawning duration model

In a recent research study (Evans-Powell et al., 2024) a model was developed for estimating mean time and duration of spawning, and how this differs with respect to female fish length. The proportion of female fish, of a given size, identified as ‘spawning capable’ (i.e. possessing ovaries that, on the basis of macroscopic and/or histological examination, are classified as mature or spawning) on any day of the year is described using a bell-shaped curve (which may or may not be symmetric). The effect of female length on spawning magnitude and duration is incorporated in the model by modifying the width and height of a ('bell-shaped') double logistic curve in accordance with fish length. The proportion of spawning capable females on a given day during the year is estimated as

$$
P=[1+\exp(-s_1(d-b_1))]^{-1}[1+\exp(s_2(d-b_2))]^{-1}H
$$

where $s_1$  and $b_1$ are the slope and inflection point for the ascending curve, respectively, $s_2$  and $b_2$ are the slope and inflection point for the descending curve, respectively, and $H$ is the maximum height of the double logistic curve. The times of year (decimal day values) corresponding to the inflection points of the ascending and descending logistic curve, for fish of total length $L$,  is calculated as

$$
b_1=\rho-\kappa_1{L}
$$
$$
b_2=\rho-\kappa_2{L}
$$
where $\rho$, $\kappa_1$ and $\kappa_2$ are model parameters. Spawning duration, for fish of length $L$, is calculated as $b_2-b_1$. The parameter $\rho$ represents the time of peak spawning, assumed to be constant across fish of all sizes. For fish of total length L, the value of $H$ is calculated as

$$
H=[1+\exp(-s_3(L-b_3))]^{-1}
$$


where $b_3$ is the inflection point and $s_3$ is the slope of the curve. This variation in height allows the maximum proportion of spawning fish, during the spawning season, to vary with fish size. The duration of the main period of spawning, for fish of total length $L$, is calculated as the difference between the (decimal day) values of $b_1$ and $b_2$.

The model is fitted using the nlminb optimiser in R (R Core Team, 2020) by minimising the overall negative log-likelihood. Setting $X_j=1$ if the $jth$ fish possessed pre-spawning or spawning ovaries, and $X_j=0$ if it did not possess such gonads, the negative log-likelihood, $\lambda$, for each model was calculated as

$$
\lambda = \sum_{j}X_j[\log_e{P_j+0.0001}+(1-X_j)\ln(1-{P_j+0.0001})]
$$

(Moore et al., 2007), where $P_j$ is the probability of the $jth$ fish being spawning capable, given its size and date of capture. The small constant 0.0001 was added to prevent natural logarithms being calculated for any calculated values of $P_j=0$.
To calculate uncertainty for both is parameters and derived variables, the fitted model parameters are assumed to conform to a multivariate normal distribution. A total of a specified number of samples of sets of values for the fitted parameters are randomly drawn from this distribution using the mvrnorm function within the MASS package in R (Venables & Ripley, 2002) and employing the variance-covariance matrix for the fitted model, calculated from the Hessian matrix produced by R when fitting the model (see Marks et al., 2021). Estimates of the lower and upper 95% confidence intervals for each of the fitted parameters and for each derived variable are calculated as the 2.5th and 97.5th quantiles of the values obtained from the sets of samples parameter estimates. Given estimates of batch fecundity and spawning frequency, the estimates of spawning duration can be used to calculate 'potential annual fecundity' for multiple spawning fish species with indeterminate fecundity. As highlighted by Evans-Powell et al. (2023) and other recent literature, use of fecundity, rather spawning biomass, may be more appropriate in some assessment models for estimating stocks status (due to issues associated with hyperallometric scaling of reproductive output with respect to fish mass). Note that this technique and associated paper is currently under review. 


### Available functions related to spawning duration model

SimulateSpawningDurationData()  
PlotSpawningDurationData()  
PlotSpawningDurationModelResults()  
CalcDailySpawnProps_SpDurMod()  
GetSpawningDurationModelResults()  



## References:

Chapman, D. G., and D. S. Robson. (1960). The analysis of a catch curve. Biometrics 16:354-368. 

Cossington, S., Hesp, S. A., Hall, N. G. & Potter, I. C. (2011). Growth and reproductive biology of the foxfish Bodianus
frenchii, a very long-lived and monandric protogynous hermaphroditic labrid. Journal of Fish Biology (2010) 77, 600–626.

deLestang, S. (2018). Could warming oceans and increased lobster biomass rates be affecting growth rates in Australia’s largest lobster fishery? Bull Mar Sci. 94, 1055–1075.

Evans-Powell, R., Hesp, S. A., Denham, A., Beckley, L. E. (2024). Implications of big, old, fat, fecund female fish (BOFFFFs) in sustaining a large demersal teleost species. Fisheries Research.

French, B., Potter, I. C., Hesp, S. A., Coulson, P. G. & Hall, N. G. (2014). Biology of the harlequin fish Othos dentex (Serranidae), with particular emphasis on sexual pattern and other reproductive characteristics. Journal of Fish Biology (2014). 84, 106-132.

Haddon, M. (2011). Modelling and Quantitative Methods in Fisheries. Chapman & Hall/CRC.

Helidoniotis F., Haddon M., Tuck G. ^ Tarbath D. (2011). The relative suitability of the von Bertalanffy, Gompertz and inverse logistic models for describing growth in blacklip abalone populations (Haliotis rubra) in Tasmania, Australia. Fisheries Research 112, 13-21.

Marks, R., Hesp, S. A., Johnston, D., Denham, A. & Loneragan, N. (2020). Temporal changes in the growth of a crustacean species, Portunus armatus, in a temperate marine embayment: evidence of density dependence. ICES Journal of Marine Science 77: 773-790.

Rogers-Bennett L., Rogers D. W. and Schultz S. A. (2007). Modelling growth and mortality of red abalone (Haliotis rufescens) in northern California. Journal of Shellfish Research, 26:719-727.

Schnute, J. (1981). A versatile growth model with statistically stable parameters. Canadian Journal of Fisheries and Aquatic Sciences 38(9) 1128-1140.

Smith, M. W., Then, A. Y., Wor, C., Ralph, G., Pollock, K. H. & Hoenig, J/. M. (2012) Recommendations for Catch-Curve Analysis, North American Journal of Fisheries Management, 32:5, 956-967.

Strain, L. W. S., Hesp, S. A., Fabris, F. and Hart, A. M. (2017). Demographic performance of Brownlip abalone: exploration of wild and cultured harvest potential. FRDC Project No. 2012/016. Fisheries Research Report No. 280, Department of Fisheries, Western Australia, 100pp.

von Bertalanffy, L. (1938). A quantitative theory of organic growth (inquiries on growth laws. II). Human Biology 10(2), 181-213.

Wakefield, C. B, Williams, A. J., Fisher, E. A., Hall, N. G., Hesp, S. A., Halafihi, T., Kaltavara, J., Vourey, E., Taylor, B. M., O'Malley, J. M., Nicol, S. J., Wise, B. W., Newman, S. J. (2020). Fisheries Research: 230. Published online October.



